A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}

### 2011-02-17 method.tex
Making method section 80 col width
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@634 ccdd05d5-56c1-461a-bf98-85f6876523b4

The detection procedure for just one set of intrinsic source parameters $\theta$ starts by {\em whitening} the measurement data stream $x$.  This involves finding a linear filter that renders the detector's noise \textsc{iid} and Gaussian.  This filter is applied to the measurement, yielding the whitened data stream $x^\mathrm W$.  The same linear filter is applied to the template $h(\theta)$, yielding the whitened template $h^\mathrm W (\theta)$.  The matched filter is the normalized cross-correlation of $h^\mathrm W$ and $x^\mathrm W$, \editorial{After introducing the $\star$ notation for cross correlation, must we define it for continuous and discrete time?}
$$
\rho(\theta) = \frac{h^\mathrm W (\theta) \star x^\mathrm W}{|h^\mathrm W (\theta)|}.
$$
This is called the signal to noise ratio, or \textsc{snr}.  The detection statistic is the modulus of this, $|\rho(\theta)|$, which has a $\chi^2$ distribution with 2 degrees of freedom in the absence of signal.%, and it indicates the probability of a signal being present \cite{wainstein:1962}\editorial{Comment about probability doesn't seem relevant.}.

@@

For concreteness and simplicity, let us consider an inspiral waveform in the quadrupole approximation, for which the time-frequency relation is
\begin{equation}
\label{eq:fgw}
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t} \right]^{3/8}.
\end{equation}
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G M_\odot / c^3 \approx 5 \mu\mathrm{s}$) and $t$ is the time relative to the coalescence of the binary~\cite{findchirppaper, kidder1992, blanchet2002, hanna2009}. \editorial{Way too many citations here.} Usually the template is truncated at some prescribed time $t = t_0$, often chosen to correspond to the innermost stable circular orbit at which $f = f_\mathrm{ISCO} \approx 4400 \, M_\odot / M$.  An inspiral signal will enter the \textsc{ligo} band at a low frequency cutoff, $f = f_\mathrm{low}$ corresponding to a time $t_\mathrm{low}$.  This template has a duration of $t_\mathrm{0} - t_\mathrm{low}$ and is critically sampled at a rate of $2 f_\mathrm{isco}$.

@@

\editorial{I'm sweeping the complex nature of the templates under the rug here.  Is it OK to regard our packing of the real and imaginary parts of the \textsc{svd} as an implementation detail?  It makes the math a little more concise.}  One may regard a bank of $N$ discretely sampled templates of length $M$ samples, $h^\mathrm W(\theta_i; t_j) = [\mathbf H]_{ij} = H_{ij}$, as a matrix.  The singular value decomposition is an exact factorization that exists for any matrix such that
\begin{equation}
\label{eq:svd-exact}
H_{ij} = [\mathbf {V \Sigma U}]_{ij} = \sum_{k=1}^{N} v_{ik} \sigma_k u_{kj},
\end{equation}
where $\mathbf V$ and $\mathbf U$ are both unitary matrices and $\mathbf \Sigma$ is a diagonal matrix.  For our purposes, we associate the rows of $\mathbf U$ with a minimal set of basis templates, which become the kernels of basis filters.  These filters give rise to the orthogonal \textsc{snr}s, $[\mathbf \rho^\perp]_k = \rho_k^\perp = \mathbf{U}_k \star x^\mathsf{W}$.  The rows of $\mathbf{V \Sigma}$ become reconstruction coefficients that map linear combinations of the orthogonal \textsc{snr}s from the basis filters back onto \textsc{snr}s for the original templates of interest: $\rho_i = \sum_k [\mathbf V \mathbf \Sigma]_{ik} \rho_k^\perp$.

In many applications of the \textsc{svd}, including ours, the matrix can be well approximated by truncating the summation in equation~(\ref{eq:svd-exact}) at $L \ll N$:
\begin{equation}
H_{ij}^\prime = \sum_{k=1}^{L} v_{ik} \sigma_k u_{kj}
\end{equation}
The cumulative sum of squares of the {\em singular values}, $\sigma_k$, measures the Frobenius norm of the approximation, such that
\begin{equation}
\frac{\| \mathbf H^\prime \|}{\| \mathbf H \|} = \left(\sum_{k=1}^{L} |\sigma_k|^2\right)^{1/2} \left(\sum_{k=1}^{N} |\sigma_k|^2\right)^{-1/2}.
\end{equation}
This is also called the \textsc{svd} tolerance.  In our application, it relates to how much \textsc{snr} is lost by discarding $N - L$ of the basis filters with the lowest singular values.

>>>>

The detection procedure for just one set of intrinsic source parameters
$\theta$ starts by {\em whitening} the measurement data stream $x$.  This
involves finding a linear filter that renders the detector's noise \textsc{iid}
and Gaussian.  This filter is applied to the measurement, yielding the whitened
data stream $x^\mathrm W$.  The same linear filter is applied to the template
$h(\theta)$, yielding the whitened template $h^\mathrm W (\theta)$.  The
matched filter is the normalized cross-correlation of $h^\mathrm W$ and
$x^\mathrm W$, \editorial{After introducing the $\star$ notation for cross
correlation, must we define it for continuous and discrete time?}
%
$$
%
\rho(\theta) = \frac{h^\mathrm W (\theta) \star x^\mathrm W}{|h^\mathrm W
(\theta)|}.
%
$$
%
This is called the signal to noise ratio, or \textsc{snr}.  The detection
statistic is the modulus of this, $|\rho(\theta)|$, which has a $\chi^2$
distribution with 2 degrees of freedom in the absence of signal.%, and it
indicates the probability of a signal being present
\cite{wainstein:1962}\editorial{Comment about probability doesn't seem
relevant.}.

@@

For concreteness and simplicity, let us consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \mu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992, blanchet2002,
hanna2009}. \editorial{Way too many citations here.} Usually the template is
truncated at some prescribed time $t = t_0$, often chosen to correspond to the
innermost stable circular orbit at which $f = f_\mathrm{ISCO} \approx 4400 \,
M_\odot / M$.  An inspiral signal will enter the \textsc{ligo} band at a low
frequency cutoff, $f = f_\mathrm{low}$ corresponding to a time
$t_\mathrm{low}$.  This template has a duration of $t_\mathrm{0} -
t_\mathrm{low}$ and is critically sampled at a rate of $2 f_\mathrm{isco}$.

@@

\editorial{I'm sweeping the complex nature of the templates under the rug here.
Is it OK to regard our packing of the real and imaginary parts of the
\textsc{svd} as an implementation detail?  It makes the math a little more
concise.}  One may regard a bank of $N$ discretely sampled templates of length
$M$ samples, $h^\mathrm W(\theta_i; t_j) = [\mathbf H]_{ij} = H_{ij}$, as a
matrix.  The singular value decomposition is an exact factorization that exists
for any matrix such that
%
\begin{equation} \label{eq:svd-exact}
%
H_{ij} = [\mathbf {V \Sigma U}]_{ij} = \sum_{k=1}^{N} v_{ik} \sigma_k u_{kj},
%
\end{equation}
%
where $\mathbf V$ and $\mathbf U$ are both unitary matrices and $\mathbf
\Sigma$ is a diagonal matrix.  For our purposes, we associate the rows of
$\mathbf U$ with a minimal set of basis templates, which become the kernels of
basis filters.  These filters give rise to the orthogonal \textsc{snr}s,
$[\mathbf \rho^\perp]_k = \rho_k^\perp = \mathbf{U}_k \star x^\mathsf{W}$.  The
rows of $\mathbf{V \Sigma}$ become reconstruction coefficients that map linear
combinations of the orthogonal \textsc{snr}s from the basis filters back onto
\textsc{snr}s for the original templates of interest: $\rho_i = \sum_k [\mathbf
V \mathbf \Sigma]_{ik} \rho_k^\perp$.

In many applications of the \textsc{svd}, including ours, the matrix can be
well approximated by truncating the summation in equation~(\ref{eq:svd-exact})
at $L \ll N$:
%
\begin{equation}
%
H_{ij}^\prime = \sum_{k=1}^{L} v_{ik} \sigma_k u_{kj}
%
\end{equation}
%
The cumulative sum of squares of the {\em singular values}, $\sigma_k$,
measures the Frobenius norm of the approximation, such that
%
\begin{equation}
%
\frac{\| \mathbf H^\prime \|}{\| \mathbf H \|} = \left(\sum_{k=1}^{L}
|\sigma_k|^2\right)^{1/2} \left(\sum_{k=1}^{N} |\sigma_k|^2\right)^{-1/2}.
%
\end{equation}
%
This is also called the \textsc{svd} tolerance.  In our application, it relates
to how much \textsc{snr} is lost by discarding $N - L$ of the basis filters
with the lowest singular values.
